Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis

### 2024-11-29 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
fix grammar error

---
title: "Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis"
subtitle: "Infant death rates rose by 0.285 per 1,000 live births in abortion-restricted states compared to protected states."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion](https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study investigates the impact of the Dobbs decision and abortion bans on U.S. infant mortality rates using 2021–2022 data. A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction. Maternal age and race are key factors, with Black mothers experiencing higher mortality rates. These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities. This study uses latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
[@tidyverse] was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` [@arrow] efficiently managed
parquet files for larger datasets. We used `modelsummary` package
[@modelsummary] to generate clean and interpretable summaries of our
regression models.We used `knitr` [@knitr] and `kableExtra`
[@kableExtra] to generate clean tables and reproducible reports, and
`usmap` for geographical visualizations. We use `here` [@here] package
simplifies the management of file paths, ensuring a reproducible
workflow by creating a consistent point of reference for file locations,
regardless of the operating system or working directory. Data cleaning
was facilitated by `janitor` [@janitor], while `lubridate` [@lubridate]
handled date-time variables. Visualization tasks were enhanced with
`ggplot2` [@ggplot2], and Bayesian regression modeling was conducted
using `rstanarm` [@rstanarm]. We use `fixest` [@fixest] to conduct the
event study. We use `caret` [@caret] to conduct the cross-validation of
the model. Together, these packages supported an efficient, reproducible
workflow for analyzing and presenting our findings.

@@

@tbl-summary shows that 50.7% of infant deaths occurred in 2022,
slightly higher than 2021 (49.3%). The majority of cases were among
mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
By race, White mothers accounted for the highest percentage (75.7%),
followed by Black or African American mothers (24.0%), and Asian mothers
(0.3%).

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
first model and 0.280 in second model), indicating a change in infant
mortality rates associated with the combination of abortion bans and the
post-injunction period.

@@

Demographic variables also exhibit some associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative
to groups 15-19 respectively. The the coefficient for Black mothers in
the second model is 6.66 relative to Asia mothers, reflecting a higher
infant mortality compared to other racial groups.

@@

In @fig-ppcheck-1 and @fig-ppcheck-2 we implement a posterior predictive
check for the first model and second model. It represents posterior
predictive checks for first model and second model, illustrating how
well each model captures the observed data. The posterior predictive
checks for first model (policy-focused) and second model (policy and
demographic factors) illustrate model fit. In @fig-ppcheck-1, the
observed ($y$) and replicated ($y_{rep}$) distributions align
moderately, reflecting general trends but limited precision.
@fig-ppcheck-2 shows improved alignment, indicating that adding
demographic factors enhances predictive accuracy and better captures the
complexities of infant mortality rates.

In @fig-posteriorvsprior1 and @fig-posteriorvsprior2 we compare the
posterior with the prior. The posterior vs. prior plots for first model
and second model illustrate the role of data in informing the posterior
estimates of the model parameters. In the first model
@fig-posteriorvsprior1, which focuses on policy-related predictors
(e.g., abortion restrictions), the data significantly shift the
posterior distributions for key variables, demonstrating their
explanatory power. second model @fig-posteriorvsprior2, incorporating
additional demographic factors such as maternal age and race, shows a
more refined and concentrated posterior distribution, emphasizing the
added predictive utility of demographic variables. The comparison
indicates that the second model provides better-informed parameter
estimates, supporting the hypothesis that demographic factors play an
essential role in understanding infant mortality in the context of
restrictive abortion laws.

@@

@tbl-priorsummary shows the prior summaries for the two models,
`first_model` and `second_model`, indicate that both use weakly
informative priors. For the intercept, a normal prior with a mean of 0
and a standard deviation of 2.5 was specified, which adjusts to a scale
of 12 after centering the predictors. The coefficients for both models
are also assigned normal priors with a mean of 0 and a scale of 2.5,
which are adjusted based on the predictors, resulting in scales such as
\[26.84, 28.43, 45.17\] for the first model and \[26.84, 28.43,
24.78,...\] for the second model. The auxiliary parameter (sigma)
follows an exponential prior with a rate of 1, adjusted to 0.2, while a
decov prior regularizes the covariance structure in both models.
Although the prior structure is similar, the second model incorporates
more predictors, leading to a larger number of coefficients with
slightly adjusted scales. These priors are designed to be weakly
informative, allowing the data to primarily guide the posterior
estimates while avoiding extreme parameter values.

@@

The RMSE values from cross-validation indicate that the first model and
second model perform similarly in predicting infant mortality rates,
with RMSE values of 5.07 and 5.06, respectively. This similarity
suggests that adding demographic factors such as maternal age and race
in the second model has only a marginal impact on predictive accuracy
compared to the first model, which primarily relies on policy-related
variables like abortion legality and timing of the injunction. An
average error of approximately 5 deaths per 1,000 live births indicates
the magnitude of prediction errors, which may or may not be acceptable
depending on the variability in the dataset. While the second model
slightly reduces the error, its added complexity may not justify the
trade-off if simplicity is a priority. However, if a nuanced
understanding of demographic influences is essential, the second model
provides a modest improvement in accuracy while offering more detailed
insights into maternal and infant health disparities. The
cross-validation process reinforces the reliability of these models by
confirming their performance across multiple data splits, reducing the
risk of overfitting.

>>>>

---
title: "Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis"
subtitle: "Infant death rates rose by 0.285 per 1,000 live births in abortion-restricted states compared to protected states."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion](https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study investigates the impact of the Dobbs decision and abortion bans on U.S. infant mortality rates using 2021–2022 data. A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction. Maternal age and race are key factors, with Black mothers experiencing higher mortality rates. These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities. This study uses the latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
[@tidyverse] was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` [@arrow] efficiently managed
parquet files for larger datasets. We used `modelsummary` package
[@modelsummary] to generate clean and interpretable summaries of our
regression models. We used `knitr` [@knitr] and `kableExtra`
[@kableExtra] to generate clean tables and reproducible reports, and
`usmap` for geographical visualizations. We use `here` [@here] package
simplifies the management of file paths, ensuring a reproducible
workflow by creating a consistent point of reference for file locations,
regardless of the operating system or working directory. Data cleaning
was facilitated by `janitor` [@janitor], while `lubridate` [@lubridate]
handled date-time variables. Visualization tasks were enhanced with
`ggplot2` [@ggplot2], and Bayesian regression modeling was conducted
using `rstanarm` [@rstanarm]. We use `fixest` [@fixest] to conduct the
event study. We use `caret` [@caret] to conduct the cross-validation of
the model. Together, these packages supported an efficient, reproducible
workflow for analyzing and presenting our findings.

@@

@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
By race, White mothers accounted for the highest percentage (75.7%),
followed by Black or African American mothers (24.0%), and Asian mothers
(0.3%).

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
first model and 0.280 in the second model), indicating a change in infant
mortality rates associated with the combination of abortion bans and the
post-injunction period.

@@

Demographic variables also exhibit some associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in the second model, relative
to groups 15-19 respectively. The coefficient for Black mothers in
the second model is 6.66 relative to Asia mothers, reflecting a higher
infant mortality compared to other racial groups.

@@

In @fig-ppcheck-1 and @fig-ppcheck-2 we implement a posterior predictive
check for the first model and second model. It represents posterior
predictive checks for the first model and second model, illustrating how
well each model captures the observed data. The posterior predictive
checks for the first model (policy-focused) and the second model (policy and
demographic factors) illustrate model fit. In @fig-ppcheck-1, the
observed ($y$) and replicated ($y_{rep}$) distributions align
moderately, reflecting general trends but limited precision.
@fig-ppcheck-2 shows improved alignment, indicating that adding
demographic factors enhances predictive accuracy and better captures the
complexities of infant mortality rates.

In @fig-posteriorvsprior1 and @fig-posteriorvsprior2 we compare the
posterior with the prior. The posterior vs. prior plots for the first model and second model illustrate the role of data in informing the posterior estimates of the model parameters. In the first model
@fig-posteriorvsprior1, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shifts the
posterior distributions for key variables, demonstrating their
explanatory power. The second model @fig-posteriorvsprior2, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the
added predictive utility of demographic variables. The comparison
indicates that the second model provides better-informed parameter
estimates, supporting the hypothesis that demographic factors play an
essential role in understanding infant mortality in the context of
restrictive abortion laws.

@@

@tbl-priorsummary shows the prior summaries for the two models, `first_model` and `second_model`, indicating that both use weakly
informative priors. For the intercept, a normal prior with a mean of 0
and a standard deviation of 2.5 was specified, which adjusts to a scale
of 12 after centering the predictors. The coefficients for both models
are also assigned normal priors with a mean of 0 and a scale of 2.5,
which are adjusted based on the predictors, resulting in scales such as
\[26.84, 28.43, 45.17\] for the first model and \[26.84, 28.43,
24.78,...\] for the second model. The auxiliary parameter (sigma)
follows an exponential prior with a rate of 1, adjusted to 0.2, while a
decov prior regularizes the covariance structure in both models.
Although the prior structure is similar, the second model incorporates
more predictors, leading to a larger number of coefficients with
slightly adjusted scales. These priors are designed to be weakly
informative, allowing the data to primarily guide the posterior
estimates while avoiding extreme parameter values.

@@

The RMSE values from cross-validation indicate that the first model and the second model perform similarly in predicting infant mortality rates,
with RMSE values of 5.07 and 5.06, respectively. This similarity
suggests that adding demographic factors such as maternal age and race
in the second model has only a marginal impact on predictive accuracy
compared to the first model, which primarily relies on policy-related
variables like abortion legality and timing of the injunction. An average error of approximately 5 deaths per 1,000 live births indicates
the magnitude of prediction errors, which may or may not be acceptable
depending on the variability in the dataset. While the second model
slightly reduces the error, its added complexity may not justify the
trade-off if simplicity is a priority. However, if a nuanced
understanding of demographic influences is essential, the second model
provides a modest improvement in accuracy while offering more detailed
insights into maternal and infant health disparities. The
cross-validation process reinforces the reliability of these models by
confirming their performance across multiple data splits, reducing the
risk of overfitting.
